---
title: "Exercises day 1, intro"
author: "ESTP Use of R in Official Statistics"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
params: 
  answer: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE
                      , eval=params$answer
                      , echo=params$answer
                      )
```
# Intro

### Exercise 1

a) Create a new R project: day 1

Execute in the `R` console the following calculations (one by one)


- $2^3$

```{r}
3*4
2^3
```


- $\frac{1}{2}+\frac{1}{2}\sqrt{5}$

```{r}
1/2 + 1/2*sqrt(5)
```


- $\frac{1+2+3+4}{4}$

```{r}
(1+2+3+4)/4
sum(1:4)/4
```

### Exercise 2

With the command `data(women)` we can load the `women` data set, which is 
available within R. 

a) Type `women` on the command-line to see the data set.
```{r}
data(women)
```

b) Copy the `women` dataset into a new variable names `ladies`.

```{r}
ladies <- women
```

c) The ladies height and weights are in inches and pounds. Add a column to the 
data with name `height_m` with height in meters (one inch is approx 2.54cm)

```{r}
ladies$height_m <- ladies$height * 2.54/100
#check
ladies$height_m
```


d) Add a column to the 
data with name `weight_kg` with weight in kilo's (one kilogram is approx $2.2046$ pounds)

```{r}
ladies$weight_kg <- ladies$weight / 2.2046
#check
ladies$weight_kg
```

e) The Body Mass Index (bmi) is calculated

$$
\frac {(\textrm{weight (kg))}}
{(\textrm{height (m)})^2}
$$ 


Calculate for the `ladies` data set the `bmi` and add this to the column.

```{r}
ladies$bmi <- ladies$weight_kg / ladies$height_m^2
ladies$bmi
```

f) Calculate

- the average (`mean`) bmi for this data set
```{r}
mean(ladies$bmi)
```

- the standard deviation for the bmi
```{r}
sd(ladies$bmi)
```

- the median for the bmi index.
```{r}
median(ladies$bmi)
```

g) Plot weight in meters against weight in kg

```{r}
plot(x = ladies$height_m, y = ladies$weight_kg, las=1)
```



```{r}
# alternative solution
plot(weight_kg ~ height_m, data = ladies, las=1)
```

### Exercise 3

`1:10` create a vector / column of numbers 1 to 10

```{r, eval=TRUE}
1:10 
```

a) Calculate

$\sum_{x=1}^{50}x^2 = 1^2 + 2^2 +\cdots +50^2$.

```{r}
sum((1:50)^2)
```


b) Generate with $:$ the vector /column $(8,4,2,1)$ and store it in `bin`.
```{r}
bin <- 2^(3:0)
```


c) Use bin to calculate the binary number $1011$.

```{r}
sum(bin * c(1,0,1,1))
```

### Exercise 4

The central limit theorem states that the sum over an infinite series of independent
random variables drawn from the same distribution, is normally distributed.
Let's test this experimentally:

a) Use the function `runif` to draw 500 random numbers from the uniform 
distribution (between 0 and 1) and
store this in the variable `x`. 

```{r}
set.seed(1)
x <- runif(500)
```

b) 
Make a histogram:

```{r}
hist(x)
```

c) Draw again 500 numbers and add these to `x`. Make a histogram and repeat until 
normally distributed.

```{r}
x <- x + runif(500)
hist(x)
```

d) **Extra** Try to make a `for` loop, repeating the process 100 times.


```{r}
for (i in 1:100){
 x <- x + runif(500)
}
hist(x, breaks=100)
```


# Reading and importing data

### Exercise 1 Csv

One of the most commonly used data format is `csv`. Let's read data in csv format.

a) Create a subdirectory "data" in your R project "day1". Copy the "adult.csv" file 
into the "data" directory. 
Try to read the csv with `read.csv`:

```{r, eval=TRUE}
adult <- read.csv("data/adult.csv")
```

b) Look at the result in the "Environment" panel. What went wrong?

_we did not use the correct import specification_

c) Try  'Import Dataset' (button upper right in Environment, or  click on "adult.csv").

d) Try 

```{r}
adult <- read.csv2("data/adult.csv")
```

e) Use `write.csv2` to save the `women` data set into csv-file "women.csv".
Tips: use `row.names=FALSE` and `na=""`.

```{r, eval = FALSE}
write.csv2("women", "./data/women.csv", row.names = FALSE, na="")
```


### Exercise 2 json

The json format is a popular format on the internet. Many websites use the format
as an intermediate transportation format between data base and web page.

JSON is considered to be human readable.

a) open de "adult.json" file and check its structure. To read json files you can use 
the `jsonlite` library.


b) Check if `jsonlite` is installed, either by searching in RStudio in "Packages" 
or by executing 

```{r}
library(jsonlite)
```


```{r, eval = FALSE}
# if not installed
install.packages("jsonlite") # install on hard disk
library(jsonlite) # load the library to make its functions available
```

c) Read "adult.json" with the function `read_json` and assign it to variable
`adult_json`. Check the data.

```{r}
adult_json <- read_json("data/adult.json")
adult_json
```


d) Read "adult.json" again and add argument `simplifyVector=TRUE` to the function. This will turn
the data into a `data.frame`. Check the result.

```{r}
adult_json <- read_json("data/adult.json", simplifyVector = TRUE)
adult_json
```


e) Write the `adults` dataset into json format and store it
as "adults_all.json". Find out what argument of the function indents the resulting json file.

```{r, eval=FALSE}
write_json(adult, "./data/adult_all.json", pretty = TRUE)
```


### Exercise 3, Sdmx

a) Install the `rsdmx` package and load it with `library(rsdmx)`

```{r}
library(rsdmx)
```


b)  The following link points to a sdmx file published by eurostat:

```{r}
sdmx_url <- "https://ec.europa.eu/eurostat/SDMX/diss-web/rest/data/nama_10_gdp/.CLV10_MEUR.B1GQ.BE/?startperiod=2005&endPeriod=2011"
```

```{r, eval = FALSE}
download.file(sdmx_url, "eurostat.xml")
```

c) Inspect the file to see what the structure is.

```{r, eval=FALSE}
# It is an xml file (sdmx)
```

d) Use the `readSDMX` function to read the file. Use the Environment / Object 
inspector to find out when this dataset was prepared and who the sender is.

```{r}
library(rsdmx)
sdmx <- readSDMX(sdmx_url)
sdmx
```


e) Use the `as.data.frame` function to transform the data into R `data.frame`

```{r}
as.data.frame(sdmx)
```

f) Which country is this data set about?

```{r, eval=FALSE}
# Belgium
```

# Summary Stats

### Exercise 1

a) Read the "adult.csv" file into variable `adults`

```{r, echo = FALSE}
adult <- read.csv2("data/adult.csv")
```

b) Calculate the `mean` age

```{r}
mean(adult$age)
```

c) Calculate the `median` hours.per-week

```{r}
median(adult$hours.per.week)
```


d) Use `summary(adult)` to get summary stats for all columns. Any peculiar findings?

```{r}
summary(adult)

# capital.gain is strange, looks like missing value coded as 99999
```



e) With `table` you can tabulate / count values:

```{r}
workclass_count <- table(adult$workclass)
workclass_count
```

f) Count the relationships in this data set

```{r}
table(adult$relationship)
```


g) Count the crossing of relations and workclass

```{r, echo = params$answer}
table(adult$relationship, adult$workclass)
```

With `tapply` you can calculate a statistic per group:

```{r, eval=FALSE, echo=TRUE}
mean_age <- 
  tapply( adult$age  # variable 
        , adult$relationship # per group
        , mean # statistic
        )
```

h) Calculate the `median` hours worked per education group
```{r}
median_hours <- 
  tapply( adult$age  # variable 
        , adult$education # per group
        , median # statistic
        )
```

i) Calculate the maximum `age` per `relationship` and `workclass` combination.

```{r}
max_age <- 
  tapply( adult$age  # variable 
        , list(adult$relationship, adult$workclass) # per group
        , max # statistic
        , na.rm=TRUE
        )
```








