---
title: "ESTP R - Reading and writing data"
output: 
  beamer_presentation:
    includes:
      in_header: ../header.tex
editor_options: 
  chunk_output_type: console
params: 
  answers: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Reading text files

### Reading

- `read.csv`, Comma separated, dot is decimal
- `read.csv2`, Semicolon separated, comma is decimal
- `read.table`, all types of text tabular format

### Writing

- `write.csv`, Comma separated dot is decimal
- `write.csv2`, Semicolon separated, comma is decimal
- `write.table`, all types of text tabular format

```{r,eval=FALSE}
dat <- read.csv("myfile.csv")
write.csv2(dat, "yourfile.csv", row.names=FALSE)
```

## File names in R

- File names are always between single or double quotes.
- File name can also be an `url`.
- Always use forward slash "/" e.g.

```{r,eval=FALSE}
dat <- read.csv("C:/users/joe/documents/foo.csv")
```

### Tip

Always work in a Rstudio project. The standard working directory of R is the project
directory. 

- use relative paths ("./")

E.g `C:/projects/foo`. Read data with `C:/projects/foo/data.csv`

```{r,eval=FALSE}
x <- read.csv("data.csv")
```

## JSON

- Java script Object Notation: very popular on websites
- Used in Web API, retrieving data via the web
- Useful for reading data from web sources
```
{"age": 34,
 "income": ["2300", "121"]
}
```

## JSON (R)

- `jsonlite` package makes it very handy to read and write json files
- allows automatic transformation from and to R data structures.

```{r, eval = FALSE, echo = TRUE}
jsonlite::toJSON(iris[1,])
```
```{r, eval = TRUE, echo = FALSE}
jsonlite::toJSON(iris[1,], pretty = TRUE)
```

## SDMX

- **Format** used by Eurostat and National Banks
- Also defines **webservice** specification
- Both data as well as metadata publication
- uses XML as storage format
```
<document>
  <age>34</age>
  <income>2300</income>
  <income>121</income>
</document>
```

## SDMX (R)

- Package `rsdmx` reads sdmx files
- Package `xml2` for generic xml files.

```{r, eval=FALSE}
library(rsdmx)
readSDMX("<link to sdmx>")
```

