---
title: "ESTP R - Descriptive statistics"
output: 
  beamer_presentation:
    includes:
      in_header: ../header.tex
editor_options: 
  chunk_output_type: console
params: 
  answers: TRUE
---
  

## Data frames

A `data.frame` is a tabular format. Technically, it is a list of vectors of the same length



```{r}
str(iris)
```



## Data frames summary

\scriptsize

```{r}
summary(iris)
```

## Handy functions

\begin{center}
\begin{tabular}{ll}
\textbf{Function}                 & \textbf{description}\\
\code{summary}                   & statistical summary \\
\code{str}                       & technical summary\\
\code{colMeans}, \code{rowMeans} & mean per column/row\\
\code{colSums}, \code{rowSums}   & sum per column/row\\
\code{names}                     & column names\\
\code{ncol} \code{nrow}          & number of columns, rows\\
\code{dim}                       & vector with \code{nrow, ncol}
\end{tabular}
\end{center}

## Selecting data


### Retrieve column with `$`

\scriptsize

```{r}
mean(iris$Sepal.Length)
# or 
# number of rows
nrow(iris)
colnames(iris)
```

##

### Retrieve rows with index

\scriptsize


```{r}
# first row (before comma)
iris[1,]

# row 2, 6 and 3 and column 1 and 2
iris[c(2,6,3),1:2]
```

##

### Row filtering

\scriptsize

```{r}
iris[iris$Species == "versicolor" & iris$Sepal.Length > 6.5, ]
```



## Plotting (1)

- The `plot` function is a generic function
- It depends on the input what kind of data is plotted
- For numeric data, a scatter plot is created


```{r, fig.height=4}
# scatter plot
plot(1:10)
```

##

```{r, fig.height=4}

# line plot
plot(1:10, type = "l")
```


## Plotting (1)

```{r, fig.height=5}
plot(weight ~ height, data=women)
```


## Plotting (2)

```{r,fig.height=5}
plot(count ~ spray, data=InsectSprays)
```

## Plotting (3)

```{r, fig.height=5}
hist(iris$Sepal.Length, breaks=20)
```


## Categorical data: `factor`

\scriptsize


A `factor` is a vector where $k$ categories have been assigned to indices $1,2,\ldots,k$.

```{r}
class(iris$Species)

# categories
levels(iris$Species)

# retrieve indices
as.integer(iris$Species)
```


## Tabulation 

\scriptsize

### Frequency table


```{r}
table(iris$Species)
```

### Cross tabulation

```{r}
iris$Sepal.Length.Class = cut(iris$Sepal.Length, breaks = 4:8)
table(iris$Species, iris$Sepal.Length.Class)
```

## Missing data

* Missing values are represented with `NA`
* `is.na()` checks for missing values
* `which()` returns indices


### Example

\scriptsize

```{r}
x <- c(3, 4, 6, NA, 3, NA)

is.na(x)

which(is.na(x))
```

## Missing data (2)

* `na.omit()` returns the vector without `NA` values

### Example

```{r}
na.omit(x)

# is the same as
x[!is.na(x)]

```

## Missing data (3)

* Most functions have a `na.rm` argument

### Example

```{r}
sum(x)

sum(x, na.rm = TRUE)
```





