---
title: "ESTP R - Basics"
output: 
  beamer_presentation:
    includes:
      in_header: ../header.tex
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 50)
```

### R is functional 

```{r}
x <- 1
x # has value of 1
```

### Always asign values to variables!

**Don't do this...**
```{r, eval=FALSE}
read.csv("test.csv")
```

**Do this**

```{r, eval=FALSE}
test <- read.csv("test.csv")
```


## Data types

Available data types: logical, numeric, integer, complex, character, and raw (not discussed)

```{r}
x <- 1
class(x)

y <- TRUE
class(y)

z <- "Some text"
class(z)
```

## Variables are vectors

Each variable is also a *vector*, i.e. a sequence of data elements of the same class:

```{r}
numbers <- 1:10
numbers

my_text <- c("Julia", "Python", "R")
my_text
```


## Vectorized operations

### Operations work on vectors

```{r}
numbers
numbers + 10
numbers^2
```

##

### Operation work on vectors (2)

```{r}
log(numbers)
mean(numbers)
paste0("My language is: ", my_text)
```


## Retrieve/set items with index

```{r}
my_text <- c("Julia", "Python", "R")
my_text[3]
my_text[3:1]
my_text[2] <- "C++"
my_text
```



## Vector generating functions

### Combine: `c`

```{r}
c(1, 5, 3, 8, 5, 3)

# Also works for vectors as input
x <- c(1, 5)
y <- c(8, 5, 3)
c(x, y)
```

##

### Repeat: `rep`

```{r}
# repeat 2, 5 times
rep(2, 5) 

# repeat vector (1,3) 5 times
rep(c(1, 3), 5) 

# repeat vector (1,3) until length output is 5
rep(c(1, 3), length.out = 5) 
```

##

### Sequence generation: `seq`

```{r}
# numbers 2 to (and including) 5
seq(2, 5) 

# 1 to 10 step size 2
seq(1, 10, by = 2) 

# 1 to 10 where output consists of 20 numbers
seq(1, 10, length.out = 20) 
```




## Comparison operators

\begin{center}

\begin{tabular}{ll}
Expression & \code{TRUE} when\\
\hline
\code{x == y} & \code{x} equal to \code{y}\\
\code{x <= y} & \code{x} less than or equal to \code{y}\\
\code{x < y}  & \code{x} less than \code{y}\\
\code{x > y}  & \code{x} greater than \code{y}\\
\code{x >= y} & \code{x} greater than or equal to \code{y}\\
\code{x != y} & \code{x} not equal \code{y}\\
\code{x \%in\% y} & \code{x} is element of \code{y}
\end{tabular}

\end{center}

## Example: \code{\%in\%}

```{r}
x <- c("Jolien", "Edwin", "John")
y <- c("Jolien", "Richard")
x %in% y
```


## Logical/binary operators

\begin{center}
\begin{tabular}{ll}
Operator  & Means\\
\hline
\code{\&} & AND\\
\code{|}  & OR (en/of)\\
\code{!}  & NOT\\
\code{all(x)} & all \code{x} equal to \code{TRUE}?\\
\code{any(x)} & at least one \code{x} equal to \code{TRUE}?
\end{tabular}
\end{center}




## Workspace

### List variables
```{r}
ls()
```

### Remove variable(s)
```{r}
rm("x")
ls()
```


## RStudio Shortcuts

* **CTRL + R** or **CTRL + Enter** Run code
* **F1** (cursor at a function name) Help
* **F2** (cursor at a function name) Go to source

