---
title: "Exercises"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
    keep_tex: no
---

```{r setup, include=FALSE, eval=TRUE}
library(dplyr)
library(tmap)
library(tmaptools)
data("World")
knitr::opts_chunk$set(echo=FALSE, eval=FALSE)
```

### Exercise 1: `tmap` 

In this exercise we'll explore thematic maps. Load the packages `tmap` and `tmaptools`. We will use the data set `World` provided by `tmap`.

a) Show a map of the world using the functions `tm_shape`, `tm_fill` and `tm_borders`. Create white borders and set the line width to 0.5.
```{r}
data("World")
tm_shape(World) + tm_fill() + tm_borders(lwd=0.5, col="white")
```
b) Create a choropleth (a colored thematic map) displaying the life expectancy of each country. Use the `virdis` color palette.
```{r}
m <- tm_shape(World) + tm_fill("life_exp", palette = viridis::viridis(5))
m
```
c) Add white country borders with 50% opacity.

```{r}
m + tm_borders("white", alpha=0.5)
```

d) Create a plot that shows the distribution of population densities. Adjust the plot to make it easier on the eye and easier and interpret.

```{r}
hist(log(World$pop_est))
```

e) Recreate the following bubble chart of estimated population sizes. Don't forget to add country borders.

```{r, eval=TRUE}
tm_shape(World) + tm_fill() + tm_borders(col="white", alpha=0.5) + tm_bubbles(size="pop_est", col="orange")
```

f) Recreate the following plot using `tm_facets`:

```{r,eval=TRUE}
tm_shape(World) +  tm_fill("pop_est_dens") + tm_facets(by="continent")
``` 

g) Create the same choropleth as in part c) of `pop_est_dens` for the Europian part of the data set. What's happening here? Explain.
```{r,eval=TRUE}
eu = World %>% filter(continent == "Europe")
tm_shape(eu) +  tm_fill("pop_est_dens") 
``` 

### Exercise 2: `sf`
In this exercise we will work with spatial data about the Netherlands which can be loaded with the command `data("NLD_muni")`.
a) 
Convert the `NLD_muni` data set to a sf data.frame and name the result `nld_sf`.
```{r}
library(sf)
data("NLD_muni")
nld_sf <- st_as_sf(NLD_muni)
```
b) Calculate the population density (inhabitants/km^2) and add this to the data set as a column named `pop_density`.

```{r}
nld_sf <- 
  nld_sf %>% 
  mutate(area = st_area(nld_sf)/1e6
        , pop_density = population / area
        )
```

d) Create a choropleth of `pop_density`

```{r}
tm_shape(nld_sf) +  tm_fill("pop_density")
```

e) Look at the distriution of population densities and assign colors to more suitable boundary values.

```{r}
tm_shape(nld_sf) +  tm_fill("pop_density", breaks = c(0, 250, 500, 750, 1000, 5000, 10000))
```

f) Create a histogram of `log(pop_density)`

```{r}
hist(log(unclass(nld_sf$pop_density)))
```

### Exercise 3: `raster`

a)
In this exercise we will use `raster`. Install and load the package and load the variables `meuse`and `meuse.grid`.

```{r}
library(raster)
data("meuse")
data("meuse.grid")
```

b)
Convert `meuse.grid` to a `SpatialPointsDataFrame` using `coordinates` from the `sp` package. Hint: look at the official R documentation of `coordinates`. Set `gridded(meuse.grid)` to `TRUE` and use the function `raster` to create an empty raster from `meuse.grid`. Examine its `summary`.

```{r}
coordinates(meuse.grid)  <- ~x+y
gridded(meuse.grid) <- TRUE
r <- raster(meuse.grid)
summary(r)
```

c)
Convert `meuse` to a `SpatialPointsDataFrame` as well using `coordinates` once more. 

```{r}
coordinates(meuse) <- ~x+y
```
d) Create a 40x40 raster using the `extent` of `meuse`.

```{r}
r40 = raster(extent(meuse), nrows = 40, ncols = 40)
```

e) Create a new raster called `rcopper` using `rasterize` and `meuse` with the amounts of `log(copper)` in the soil.

```{r}
rcopper = rasterize(meuse, r40, "copper")
```

f) Repeat the previous exercise for lead concentration and name the new raster `rlead`.

```{r}
rlead = rasterize(meuse, r40, "lead")
```

g) Combine the two rasters with `stack` and plot this variable to obtain two plots side-by-side.

```{r}
plot(stack(rcopper, rlead))
```

We will finish this subject with some exercises about a different built-in data set: `volcano`. This data set gives topographic information for the Maunga Whau volcano in New Zealand on a 10m by 10m grid.
h) Create a raster `v` out of `volcano` and plot `v`. Use the `viridis` color palette.

```{r}
v <- raster(volcano)
plot(v, col = viridis::viridis(255))
```
i)  Create a contour plot for this raster.

```{r}
contour(v)
```

j) Create a `filledContour` plot of `v` and limit the number of colours to 10.

```{r}
filledContour(v, nlevels = 10)
```

k) Recreate the following 3D surface plot of `v` using `persp`:

```{r, eval=TRUE}
library(raster)
v <- raster(volcano)
persp(v, border="#00000044")
```

### Exercise 4: `leaflet`

The `leaflet` package uses so-called lat/lon coordinates (wgs84 or "longlat").
We will use the data set of the Netherlands we created in exercise 2 (`nld_sf`) which uses a different (Dutch) projection called "Rijksdriehoeksstelsel". 
a) Convert the projection to wgs84 using `st_transform`.

```{r}
nl <- st_transform(nld_sf, get_proj4("longlat")) 
```

b) Create an interactive choropleth in leaflet of population sizes. First create a `colorNumeric` function and then use this as an argument in `addPolygons` to fill the colors of the polygons.

```{r}
library(leaflet)

nl$pop_density = unclass(nl$pop_density)
pal <- colorNumeric("viridis", range(nl$pop_density))

leaflet(nl) %>% addTiles() %>%  addPolygons(fillColor=~pal(pop_density))
```
c) The library `tmap` also provides an easy-to-use function to create interactive leaflet maps called `tmap_leaflet`. Redo exercise 2e but this time make the choropleth interactive.

```{r}
m = tm_shape(nld_sf) +  tm_fill("pop_density", breaks = c(0, 250, 500, 750, 1000, 5000, 10000))
tmap_leaflet(m)

```

d) Create a bubble chart of `World` showing the number of inhabitants for each country. Explain what happened to France.

```{r}
b <- tm_shape(World) + tm_fill() + tm_bubbles("pop_est")
tmap_leaflet(b)
```
