---
title: "Exercises Shiny"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
params:
  answers: no
editor_options: 
  chunk_output_type: console
---
  
```{r, include=FALSE}
knitr::opts_chunk$set( echo = params$answers
                       , eval = params$answers
                       , comments = NA
                       , message=FALSE
                       , warning = FALSE
)
library(shiny)
```

In this session, we will create a dashboard to analyse the `Wage` dataset:

```{r,eval=TRUE,echo=TRUE}
Wage <- read.csv("data/Wage.csv")
str(Wage)
```

a) Create a shiny app that shows a histogram of the column `wage`, with a slider for the number of bins

```{r,eval=FALSE}
ui <- fluidPage(
  titlePanel("Wage data"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Number of bins", min = 10, max = 100, step  = 5, value = 30)
    ),
    mainPanel(
      plotOutput("hist")
    )
  )
)


server <- function(input, output) {
  output$hist <- renderPlot({
    hist(Wage$wage, breaks = input$n, main = "Histogram of wage", xlab = "Wage")
  })
}

shinyApp(ui, server)
```

##

b) We would like to create histograms for different subgroups in the data. We can group the data according to any categorical data variable in the data. For instance, for `education`, we can create the subgroups: `r sort(unique(Wage$education))`. Create a shiny app where the user can select the grouping variable (use `selectInput`). Use `ggplot2` to render histograms, one for each category (ideally stacked vertically). The variable `age` is an integer, but we can create age-groups with the function `cut`:
  
```{r,eval=TRUE,echo=TRUE}
Wage$age_class = cut(Wage$age, breaks = c(18, 35, 55, 80), right = FALSE, include.lowest = TRUE)
```


```{r,eval=FALSE}
library(ggplot2)

ui <- fluidPage(
  titlePanel("Wage data: subgroups"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Number of bins", min = 10, max = 100, step  = 5, value = 30),
      selectInput("grp", "Grouping variable", choices = c("age_class", "maritl", "race", "education", "region", "jobclass", "health", "health_ins"), selected = "maritl")
    ),
    mainPanel(
      plotOutput("hist")
    )
  )
)


server <- function(input, output) {
  output$hist <- renderPlot({
    ggplot(Wage, aes(x=wage)) + geom_histogram(bins = input$n) + facet_grid(rows = input$grp)
  })
}

shinyApp(ui, server)
```

