---
title: "Project management"
author: "ESTP Use of R in Official Statistics"
output: 
  beamer_presentation:
    includes:
      in_header: ../header.tex
    fig_caption: false
params: 
  answers: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Project management

### The Statistical Value Chain

Define a number of 'half-products' (data sets) with set quality demands.

\begin{picture}(0,0)
\put(0,-75){
\includegraphics[height=0.30\textheight]{fig/fig04_value-chain.png}
}
\end{picture}


## Project management

### Value chain
- **Raw data**: as it enters our offices
- **Input (tidy) data**: records and variables identifiable
- **Valid data**: content is checked
- **Statistics**: aggregate
- **Output**: reports, plots, ...
  
### Directory structure

- Create a directory for each half-product
- The R code in each directory creates the data in that directory
  - The R code can be in a rmarkdown document

# R packages and the R help system

## R packages

### A R package consists of:

- New functions and/or data with 
    - documentation (**obligatory**)
    - examples (optional) 
    - automatic tests (optional)
    - vignettes ($=$tutorials, optional)

### Publication

Via a sort app store:

- Centrally via **`https://cran.r-project.org`**
- Or see via  **`https://r-pkg.org/`** 

### Installation

```{r,eval=FALSE}
install.packages("packagename")
```

## What package should I use?

### The internet helps them who help them selves

1. Google your application (Use `R` as keyword);
2. read blogs (`r-bloggers`), SO questions;
2. read vignettes, but mainly:
4. **Install and get started**

## R documentation

### Built-in help
- Standard technical reference, (often) with links, author, examples. 
- Meant to document a function independent of other functions.

### Vignettes
- Tutorials, walkthroughs
- Task orient, meant to introduce a user to the package.


