---
title: "Statistical modeling"
author: "ESTP Use of R in Official Statistics"
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
fig_caption: false
editor_options: 
  chunk_output_type: console
---

```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment=NA)
library(tidyverse)
```

## Contents

- Why statistical modeling?
- Inference
- Predict

## Official Statistics

Statistics can be:

- _descriptive_: describing/summarizing the "state"
- _predictive_: estimate an unknown/incomplete parameter/indicator
- _forecasting_: predicting the future value of indicators.
- _inferential_: find out the mechanism, contributing parts.

_Prediction aka Machine Learning_

## Official Statistics

- By origin Official Statistics is **descriptive**: it describe the "state" 
(of the state :-) )

- But **observations are never complete**. A survey is a method
to do a small (incomplete) measurement to calculate statistics on (groups in) the
population: _estimation is a prediction method_. 

- But **observations can have errors and missing values**. Fixing values is a form of
_prediction_. 

- But analysing time series, finding trend and seasonal patterns is very similar
to _forecasting_: often it is _now-casting_.

## Prediction

- survey estimation predicts for groups: totals (and or) means.
- most prediction works on observations.
- can be used together: _small area estimation_ combines survey and prediction methods

## Statistical models

### Statistical models are:

"smart", sensible guessing machines, calibrated with known data:
  
  > If we measure the happyness of 0.1% of the population (smartly selected), this
  will (probably)  hold for the whole population.
  
  > If we do not know the NACE code for this enterprise, but it has similar properties
  as whole-trade enterprises, it (probably) is a whole-trade enterprise.
  
  
## Descriptive stats

- Can be done easily with `dplyr` and R: 
- Counting / tabulating: `table`, `dplyr::count`, `dplyr::summarize`, `dplyr::n()`
- variables: `mean`, `median`, `sum`
- per group: `dplyr::group_by` (or `tapply`)


## Prediction

- For surveys the `survey` package is of interest.
- For now we focus on predicting values for individual records:

### Useful in official stats for

- classification (e.g. enterprises, economic activity)
- imputation for missing or erroneous values. 
- estimating statistics for small groups (small area estimation)
- clustering (technically not prediction, but can reveal structure in data that
can be used to predict which cluster an observation belongs to)

## Statistical prediction

### Model
  $$
  Y = f(X) + \varepsilon
  $$

- $Y$: Predicted variable
- $f$: Relation between $Y$ and $X$
- $X=(X_1,X_2,\ldots,X_p)$: predictive variables (predictors), aka independent variables
- $\varepsilon$: Noise, independent of $X$ en $Y$. 


## Prediction

### Estimated model
$$
\hat{Y} = \hat{f}(X)
$$
\begin{itemize}
  \item $\hat{Y}$ the estimated value for  $Y$.
  \item $\hat{f}$ the estimated model.
\end{itemize}

### Remark
For individual observations
$$
y_i = \hat{f}(x_i),
$$
the relation $f$ works as a \emph{black box} / data generating machine:
input $x$ generates value $y$.

## Example

```{r,message=FALSE,echo=FALSE}
library(ISLR)
plot(jitter(wage) ~ jitter(age), data=Wage
     , xlab="Age"
     , ylab="Income"
     , main="Income (thousands dollars)"
     , las=1
     , cex.axis=1.2
     , cex.lab=1.2
     , cex.main=1.5
     , pch=16
     , col=adjustcolor("black",alpha.f=0.6
  ))
pred <- loess.smooth(Wage$age,Wage$wage)
lines(pred,lwd=8,col='blue')
pred1 <- lm(wage ~ age, data=Wage)
lines(Wage$age, predict(pred1),col='red',lwd=8)
```

$X$: age, $Y$: income, \textcolor{red}{$f$}: lin. reg., \textcolor{blue}{$f$}: Loess.

## Prediction error (classical)

### Estimated prediction error
$$
\textsf{MSE}(\hat{f}) = \frac{1}{n}\sum_{i=1}^n(y_i - \hat{y}_i)^2
$$
\begin{itemize}
\item $y_i$: Observed values
\item $\hat{y}_i$: Predicted values $\hat{y}_i=\hat{f}(x_i)$
\item $\textsf{MSE}$ Mean Square Error for estimated model $\hat{f}$.
\end{itemize}

## Prediction

### Numerical $Y$

- e.g. `turnover`
- prediction is called **regression**

### Categorical $Y$

- e.g. `nace`
- prediction is called **classification**

## Prediction methods

- linear models: `lm`
- generalized linear models: `glm`
- decision trees: `rpart` (from R package `rpart`)
- randomForests: `randomForest` (from R package `randomForest`)
- etc.

## Model in R

### Formula : 

```{r, eval =FALSE}
wage ~  age + education 
# wage depends on age and education
```

```{r, eval = FALSE}
model <- lm(wage ~ age, data = Wage) 
# create a linear model with this 
```

## Linear model in R

```{r, eval = TRUE}
model <- lm(wage ~ age, data = Wage) 
# create a linear model with this 
model
```

## Linear model in R

\tiny
```{r, eval = TRUE}
summary(model)
```



