---
title: "Graphical User Interfaces and dashboards with shiny"
author: "ESTP Use of R in Official Statistics"
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
fig_caption: false
editor_options: 
  chunk_output_type: console
---

## What is `shiny`?

- `shiny` is an R package to create web-based applications
- It is also popular outside the R community, where it is refered to as "R Shiny"
- An application created with Shiny is often called a "Shiny App" or simply an "app".
- Applications can run in the browser, so no additional software required
- R is required, either local or via a server

## Web development

- `shiny` uses HTML/CSS/Javascript under the hood, the main languages for professional web development
- No knowledge of HTML/CSS/Javascript required to create shiny apps, but...
- additional HTML/CSS/Javascript make shiny fully customizable and extendable

## `ui` and `server`

A `shiny` app consists of two main functions:

1. A `ui` which stands for user interface: this contains all visual components, such as buttons, sliders, and (placeholders for) interactive charts and tables
2. A `server` which contains the R back-end for creating interaction between the visual components. E.g., when a slider is changed, the histogram should be updated.

## Example: ui

\scriptsize

```{r, eval=FALSE}
library(shiny)
ui <- fluidPage(
  titlePanel("Histrogram of random data"),
  sidebarPanel(
    sliderInput("k",
               "Number of observations",
               min = 100,
               max = 10000,
               value = 1000,
               step = 100),
    radioButtons("type",
                 "Distribution type",
                 choices = c("Uniform", "Normal", "Log-normal"),
                 selected = "Uniform")),
  mainPanel(
    plotOutput("hist")
  )
)
```

## Example: server


```{r, eval=FALSE}
server <- function(input, output) {
  output$hist <- renderPlot({
    if (input$type == "Uniform") {
      hist(runif(input$k))
    } else if (input$type == "Normal") {
      hist(rnorm(input$k))
    } else if (input$type == "Log-normal") {
      hist(rlnorm(input$k))
    }
  })
}
```

## Example: run app

```{r, eval=FALSE}
shinyApp(ui, server)
``` 


## Example: server (alternative)

```{r, eval=FALSE}
server <- function(input, output) {
  
  get_data <- reactive({
    type <- input$type
    k <- input$k
    
    fun = switch(type, Uniform = runif, Normal = rnorm, 'Log-normal' = rlnorm)
    
    fun(k)
  })
  
  output$hist <- renderPlot({
    data <- get_data()
    hist(data)
  })
}
```


## Page layout functions

### Page functions 
- All UI items should ultimately be contained in a `page`
- Standard page functions are `fluidPage`, `fixedPage`, `fillPage` and `navbarPage`


### Layout functions
- Layout functions organize the UI items
- Many layouts available, e.g. `sidebarLayout`, `verticalLayout`, `splitLayout`
- See (and run) examples of these functions to see the layouts


## Control widgets

Standard shiny widgets:

\scriptsize

- actionButton	Action Button
- checkboxGroupInput	A group of check boxes
- checkboxInput	A single check box
- dateInput	A calendar to aid date selection
- dateRangeInput	A pair of calendars for selecting a date range
- fileInput	A file upload control wizard
- helpText	Help text that can be added to an input form
- numericInput	A field to enter numbers
- radioButtons	A set of radio buttons
- selectInput	A box with choices to select from
- sliderInput	A slider bar
- submitButton	A submit button
- textInput	A field to enter text


## HTML elements

HTML elements can be added via the `tags` list:

```{r,eval=FALSE}
ui <- fluidPage(
  tags$h1("Heading 1"),
  tags$h2("Heading 2"),
  tags$p("Paragraph"),
  tags$br(),
  tags$a(href = "https://shiny.rstudio.com", "Shiny!"))
server <- function(input, output){}
shinyApp(ui, server)
```



## Output widgets

Always consist of two functions: a placeholder in the UI and a rendering function in the server

Standard output widgets are:

- `plotOutput` and `renderPlot`
- `tableOutput` and `renderTable`
- `textOutput` and `renderText`
- `uiOutput`/`htmlOutput` and `renderUI`

The latter is used to update the control widgets


## Reactive and observe

Shiny contains two families of interactive functions:

1. **reactive** These functions (`reactive`, `reactiveValues`, ...) are lazy: they are only executed when they are called elsewhere
2. **observe** These functions (`observe` and `observeEvent`) are eager: they are continually monitoring for changes in the dependencies



## Hosting shiny apps

Shiny apps can be runned locally, or hosted online, e.g. at https://www.shinyapps.io


