---
title: "The UN dataset"
author: "Me, myself and I"
date: "March 5, 2021"
output:
  pdf_document: default
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r}
un <- read.table("data/UnitedNations.txt", header=TRUE)
```

( _italic_ texts should be generated with an r statement
)

### Introduction

In this report we revisit the dataset of the United Nations. 
This dataset has _`r nrow(un)`_ rows. Each row describes variables of a country. 
The countries are subdivided into _`r length(unique(un$region))`_ regions.
The following table shows the number of countries per region.

```{r, results='asis'}
out <- un %>%
  group_by(region) %>%
  summarize(countries=n())
pander::pandoc.table(out)
```

### Life expectance and GDP

The life expectency and the Gross Domestic Product are important indicators
for well being. In the following table are the average life expectancy per country of men
women.

```{r, results='asis'}
un %>%
  group_by(region) %>%
  summarize(men = mean(lifeMale,na.rm=TRUE)
            ,  women=mean(lifeFemale,na.rm=TRUE)) %>%
  pander::pandoc.table("Mean life expectancy (years)")
```

```{r}
lowest_women <- un %>% 
  filter(lifeFemale == min(lifeFemale, na.rm=TRUE))
```

The country with the lowest life expectancy for women is _`r lowest_women$country`_
with a life expectancy of _`r lowest_women$lifeFemale`_ year.


The following barchart shows the GDP per capita per region:

```{r}
un %>% group_by(region) %>% 
  summarize(gdp = mean(GDPperCapita,na.rm=TRUE))  %>%
  ggplot(aes(x=region, y=gdp, fill=region)) +
  geom_col(show.legend = FALSE) +
  labs(title="GDP per capita in $", x="regio", y="GDP")
```


### Relation between life expectancy and GDP

It is interesting to see if the GDP per capita relates to life expectancy. Let's
look at this relationship for women.

```{r,message=FALSE, warning=FALSE}
un %>% ggplot(aes(x=GDPperCapita, y=lifeFemale, color=region)) +
  geom_point() + scale_x_log10() + geom_smooth(method='lm') + 
  labs(title="Life expentancy of women") + theme_bw()
```




