---
title: "Exercises Imputation"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
params:
  answers: no
---

```{r, include=FALSE}
knitr::opts_chunk$set( echo = params$answers
                     , eval = params$answers
                     , comments = NA
                     , message=FALSE
                     , warning = FALSE
                     )
library(tidyverse)
library(simputation)
```


### Exercise 1, imputation

a) Install R package `simputation` (if you have not done so)

```{r, eval=FALSE, echo=TRUE}
install.packages("simputation", dependencies = TRUE)
```

b) load the data set "data/iris_missing.csv". 
Find out on which colums data is missing. 

```{r}
iris_missing <- read.csv("data/iris_missing.csv")
summary(iris_missing)
```

c) Find out which on rows data is missing (tip `complete.cases()`)

```{r}
which(!complete.cases(iris_missing))
```


c) Calculate the mean of `Sepal.Length`.

```{r}
mean(iris_missing$Sepal.Length, na.rm=TRUE)
```


d) Load the simputation library and impute `Sepal.Length` with a linear model dependent
on `Sepal.Width` and `Species` (use `impute_lm`). Store the new data set in 
`iris_imputed1`

```{r}
library(simputation)
iris_imputed1 <- impute_lm(iris_missing, Sepal.Length ~ Sepal.Width + Species)
```


e) Calculate the mean of `iris_imputed1$Sepal.Length`. Any ideas why there are still `NA`s? Which rows have not been imputed?

```{r}
mean(iris_imputed1$Sepal.Length)
mean(iris_imputed1$Sepal.Length, na.rm=TRUE)

incomplete <- !complete.cases(iris_imputed1)
# wel Sepal.Length is missing
iris_imputed1[incomplete, ]
```


f) Impute remaining NA values of  `Sepal.Length` in `iris_imputed1` with `impute_median` per `Species` (i.e. `Sepal.Length ~ Species`) and store the result in `iris_imputed2`. 
What is the resulting `mean`?

```{r}
iris_imputed2 <- impute_median(iris_imputed1, Sepal.Length ~ Species)
mean(iris_imputed2$Sepal.Length)
```


g) Impute `Sepal.Width` with decision tree model (`cart`), using all other variables.
And store this in `iris_imputed3`.

```{r}
iris_imputed3 <- impute_cart(iris_imputed2, Sepal.Width ~ .)
```


h) Use the `%>%` operator to chain to create an imputation pipeline.

```{r}
iris_imputed <- 
  iris_missing %>% 
  impute_lm(., Sepal.Length ~ Sepal.Width + Species) %>% 
  impute_median(., Sepal.Length ~ Species) %>% 
  impute_cart(., Sepal.Width ~ .)
```


### Exercise 2

a) Read `data/companies.csv`

```{r}
companies <- read.csv("data/companies.csv")
```


b) Make a separete data frame, selecting columns 7--14 (`staff`--`vat`)

c) Implement the following imputation sequence:
    - Impute `turnover` by copying the `vat` variable (`impute_proxy`)
    
```{r}
```
    
    - Impute `staff` with a robust linear model based on `staff.costs`
    
```{r}

```
  
    - Impute `staff` with a robust linear model based on `total.costs`
    - Impute `profit` as `total.rev - total.costs` (`impute_proxy`)
    - Impute everything else using `missForest` (formula: `. ~ .`)
```{r}
library(tidyverse)
library(missForest)
companies %>% 
  dplyr::select(., staff:vat) %>% 
  impute_proxy(., turnover ~ vat) %>% 
  impute_rlm(staff ~ staff.costs) %>% 
  impute_rlm(staff ~ total.costs) %>% 
  impute_proxy(staff ~ total.rev - total.costs) %>% 
  impute_mf(., . ~ .)
```
