---
title: "Exercises StatModel"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
params:
  answers: TRUE
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set( echo = params$answers
                     , eval = params$answers
                     , comments = NA
                     , message=FALSE
                     , warning = FALSE
                     )
library(tidyverse)
```
We will fit two different models to the "data/Wage.csv" dataset. A linear model and a Decision Tree model. 

(The proper way of comparing models is to use cross validation, e.g. splitting your data in a test and train set, but to simplify things, we will neglect this.)

a) Load the "data/Wage.csv" file into variable `Wage`. Make a summary. 
We will build a model that tries to predict `wage` from the other variables.

```{r}
Wage <- read.csv("data/Wage.csv")

```

b) Fit the following linear model (lm):

```{r, echo = TRUE}
model_lm1 <- lm(wage ~ age, data = Wage)

model_lm1
summary(model_lm1)
model_lm1$coefficients

```
Print the `model_lm`, `summary` and coefficients `coef`.

c) Get the predictions by using R function `predict`

```{r, echo = TRUE}
wage_lm1 <- predict(model_lm1)
```

We will look at the root mean squared error (rmse) as a measure to compare models.

```{r, echo = TRUE}
rmse_lm1 <- sqrt(mean((Wage$wage - wage_lm1)^2))
rmse_lm1

```

d) Use predict to find out what wage is predicted for `age = 35`. Use argument `newdata` with a `data.frame(age = 35)`

```{r}
predict(model_lm1, newdata = data.frame(age = 35))
```


e) repeat b and c for `wage ~ age + education`. Is this a better model fit?

```{r}
model_lm2 <- lm(wage ~ age + education, data = Wage)
wage_lm2 <- predict(model_lm2)
rmse_lm2 <- sqrt(mean((Wage$wage - wage_lm2)^2))
rmse_lm2

# Another option to compare results:
anova(model_lm1, model_lm2)

```


e) With the function `rpart` a decision tree (CART) can be used to predict

```{r, echo = TRUE}
library(rpart)
model_cart1 <- rpart(wage ~ age, data = Wage)
```

Calculate the mse for this model and store it in `mse_cart1`

```{r}
wage_cart1 <- predict(model_cart1)
```

e) Repeat this for `wage ~ age + education` and predict this with you own age and education for both models.

f) What is the `mse` for both models?

### Exercise 2, Classification

a) Load the dataset "data/Caravan.csv". We will try to predict whether a person 
has purchased a caravan.


```{r}
Caravan <- read.csv("data/Caravan.csv")

names(Caravan)
summary(Caravan)
```

b) Fit a cart model with `rpart` to predict a purchase. Use all variables.

```{r}

model <- rpart(Purchase ~ ., data= Caravan)

```

c) Predict the labels using the following statement:

```{r, echo =TRUE}
purchase_cart <- predict(model, type = "class")
```

d) create cross table where you compare how well the prediction is.

```{r}
tab <- table(predicted = purchase_cart, observed = Caravan$Purchase)
tab

```

e) what is the precision and recall?

```{r}

# TP/(TP+FP) = precision
# TP/(TP+FN) = recall

TP <- tab[2,2]
FP <- tab[2,1]
FN <- tab[1,2]

TP/(TP+FP)
TP/(TP+FN)


prop.table(tab, 1)
prop.table(tab, 2)
 
```









