---
title: "Exercises day 2, tidyverse"
author: "ESTP Use of R in Official Statistics"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
params:
  answers: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( eval=params$answers
                     , echo=params$answers
                     )
```


Load the R package `tidyverse` with the command `library(tidyverse)`

### Exercise 1
We will look at the `titanic` dataset. Load the file `./data/titanic.csv` and
save it in the variable `titanic`. This file contains the passengers of the Titanic.

a) How many passengers are listed in this file? Calculate this with `summarize`.

```{r}
library(tidyverse)
titanic <- read_csv2("./data/titanic.csv")

count(titanic)
summarize(titanic, n  = n())
nrow(titanic)
```

b) How many passengers have survived? Also use a construction with `group_by`

```{r}
titanic %>%  
  filter(., survived == "yes") %>%
  count(.)

count(titanic, survived)

titanic %>% group_by(survived) %>% summarise(n = n())
```

c) Calculate the mean age of the passengers. What was the mean age of the women
on board? Calculate this with `summarize`

```{r}
mean(titanic$age, na.rm=TRUE)

titanic %>% 
  summarize(mean_age = mean(age, na.rm=TRUE))

titanic %>% 
  filter(sex == "female") %>% 
  summarize(mean_age = mean(age, na.rm=TRUE))

titanic %>% 
  group_by(sex) %>% 
  summarize(mean_age = mean(age, na.rm=TRUE))
```

d) Calculate the mean age of the passengers, by `sex` and `survived`.

```{r}
titanic %>% 
  group_by(sex, survived) %>% 
  summarize( mean_age = mean(age, na.rm=TRUE))
```

e) As in d), but now also the min and max age for each group. 

```{r}
titanic %>% 
  group_by(sex, survived) %>% 
  summarize( mean_age = mean(age, na.rm=TRUE)
           , max_age = max(age, na.rm=TRUE)
           , min_age = min(age, na.rm=TRUE)
           )
```


### Excercise 2

Create a data pipeline that does the following with the titanic data set

a) Select the men.
b) Group on passenger class
c) Calculate the number of survivers (tip: you can sum over the number of `TRUE`)

```{r}
titanic %>% 
  filter(sex == "male") %>% 
  group_by(passengerClass) %>% 
  summarize(n_survived = sum(survived == "yes"))

titanic %>% 
  filter(., sex == "male", survived == "yes") %>% 
  count(., passengerClass)
```

d) As in a-c, but now the percentage survivors per passengerclass

```{r}
titanic %>%
  filter(sex=="male") %>%
  group_by(passengerClass) %>%
  summarise( survival_rate = mean(survived == "yes")
           , survived_percentage = 100 * survival_rate
           )
```

### Exercise 3:

We continue with the `Titanic` dataset

a) Use the function `table` to count the number of passengers who survived and did
not survive.

```{r, eval=FALSE, echo=params$answers}
table(titanic$survived)
```

b) Use the function `table` to cross tabulate the number of passengers to `survived`
and `sex`. Store the result in the variable `survivors`. Plot the variable `survivors`.

`survivors`. Plot de variable `survivors`.
```{r, eval=FALSE, echo=params$answers}
survivors <- table(
  titanic$survived
  , titanic$sex)
plot(survivors)

```

c) Execute the following commands and interpret the result(s).

- `prop.table(survivors)`
- `prop.table(survivors,1)`
- `prop.table(survivors,2)`

```{r, eval=FALSE, echo=params$answers}
prop.table(survivors)    # fractions, sum() to 1
prop.table(survivors, 1) # row totals are equal to 1
prop.table(survivors, 2) # column totals are equal to 1
```

d. Same as before but with  `addmargins`.
```{r, eval=FALSE, echo=params$answers}
dplyr::
addmargins(survivors)    # all totals
addmargins(survivors, 1) # column totals
addmargins(survivors, 2) # row totals
```

e) We derive an `ageClass` variable, use the command:
```{r,echo=TRUE}
titanic$ageClass <- cut(titanic$age
   , breaks=c(0,17,64,Inf)
   , labels=c("young","middle","old")
) 
```

Create a table with `ageClass` vs `passengerClass` and answers the following questions:

- How many passengers were on board of the titanic
- How many passengers were counted in the last table: explain

```{r, eval=FALSE, echo=params$answers}
tab <- table(titanic$ageClass, titanic$passengerClass)
#compare
sum(tab)
# with 
nrow(titanic)
# use
table(titanic$ageClass, titanic$passengerClass, useNA = "ifany")
# to see that many age are not available
```


### Exercise 4

a. Read the files `addressdata.csv` and `addressinfo.csv` with R

Name the variables in R  `address` and  `addressinfo`. (check the results!)

```{r, eval=FALSE, echo=params$answers}
address <- read.csv2("data/addressdata.csv", stringsAsFactors = FALSE)
addressinfo <- read.csv2("data/addressinfo.csv", stringsAsFactors = FALSE)
```
b. Use `inner_join` to enrich `address` with information from `addressinfo`.
Match on street and place name.

```{r, eval=FALSE, echo=params$answers}
dat <- inner_join(address, addressinfo, by = c("street" = "streetname", "place"))
```
c. Do the same with now with `left_join`. What is the difference?

```{r, eval=FALSE, echo=params$answers}
dat <- left_join(address, addressinfo, by = c("street" = "streetname", "place"))
```

d. Select all records in `address` that cannot be matched met addressinfo with  `anti_join`.
```{r,eval=FALSE, echo=params$answers}
anti_join(address, addressinfo, by = c("street" = "streetname", "place"))
```


### Exercise 5, Missing en special values

Predict the outcome of the following R statements

a. `mean(c(1, 1, NA, 1))`
```{r,eval=FALSE, echo=params$answers}
# NA, because the mean depends also on the unknown values
```
b. `length(c(1,2,NA,4))`
```{r,eval=FALSE, echo=params$answers}
# 4, NA fills a spot 
```
c. `mean(c(1,4,4,NA,NULL), na.rm=TRUE)`
```{r,eval=FALSE, echo=params$answers}
# (1 + 4 + 4)/3  = 3. NULL does not add an element and NA is removed with
# na.rm=TRUE.
```
d. `c(1,2,NA) == NA`
```{r,eval=FALSE, echo=params$answers}
# NA NA NA. Each comparison with NA is NA: unknown does not equal unknown
```
e. `is.na(c(1,2,NA))`
```{r,eval=FALSE, echo=params$answers}
# FALSE FALSE TRUE
```
f. `sum(is.na(c(1,2,NA)))`
```{r,eval=FALSE, echo=params$answers}
# 1
```



### Exercise 6. Reading files with `read_csv`

The file `cars.csv` contains 12 columns:

a. Read the file with \code{read\_csv2}.
```{r,eval=FALSE, echo=params$answers}
cars <- read_csv2("data/cars.csv")
```
b. Read the file with \code{read\_csv2} and specify the correct column types. 
You may read \code{model}, \code{engine} and \code{gearbox} as text variables.

```{r,eval=FALSE, echo=params$answers}
cars1 <- read_csv2("data/cars.csv", col_types="cdccdddddddd")
```
c. Read the file with \code{read\_csv2} and make sure that  \code{engine} en
\code{gearbox} are factor variables
```{r,eval=FALSE, echo=params$answers}
cars2 <- read_csv2("data/cars.csv"
    , col_types = list(
        engine  = col_factor(c("V", "straight"))
      , gearbox = col_factor(c("manual", "auto"))
    )
)
```
d. Read the file with \code{read\_csv2} and skip column 3 and 5
```{r,eval=FALSE, echo=params$answers}
cars3 <- read_csv2("data/cars.csv", col_types="cd-c-ddddddd")
```
  
### Exercise 7

a. Read the dataset `adult.csv`. In this file missing values are stored with as " ?".
(<space> <question mark>). Make sure that these are read in as `NA` (see help of
read_csv2).

```{r,echo=params$answers}
library(readr)
adult <- read_csv2("data/adult.csv", na="?", trim_ws = TRUE)
```

b. Create a frequency table of age, what is the modal age?

```{r,eval=FALSE, echo=params$answers}
ages <- table(adult$age)
ages[ages==max(ages)]
# or
which.max(ages)
sort(ages, decreasing=TRUE)[1]
```

c. Cross tabulate `workclass` and `income.class` (with `count` or with `table`) 
(`workclass`, `income.class`)
```{r,eval=FALSE, echo=params$answers}
tab <- table(adult$workclass, adult$income.class)
```


d. Which `native.country` has the largest female/male ratio?

```{r,eval=FALSE, echo=params$answers}
tab <- table(adult$native.country,adult$sex)
ratio <- tab[,"Female"]/tab[, "Male"]
which.max(ratio)
tab["Holand-Netherlands",]
```




