---
title: "Exercises day 2, data visualization"
author: "ESTP Use of R in Official Statistics"
output: 
  pdf_document:
    includes:
      in_header: ../exercise.tex
params:
  answers: TRUE
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set( echo = params$answers
                     , eval = params$answers
                     , comments = NA
                     , message=FALSE
                     , warning = FALSE
                     )
library(tidyverse)
```

### Exercise 1

a)  Load the data set mtcars and the library `ggplot2`

```{r, eval = TRUE, echo = TRUE}
data(mtcars)
library(ggplot2)
```

b)  Create a scatter plot with `wt` on the x-axis and `disp` on the y-axis.

```{r}
ggplot(mtcars, aes(x = wt, y = disp)) + geom_point()
```

c)  Make the points blue.

```{r}
ggplot(mtcars, aes(x = wt, y = disp)) + geom_point(color="blue")

# in case the color is defined in ggplot, make sure to use the I function:
# ggplot(mtcars, aes(x = wt, y = disp, color = I("blue"))) + geom_point()
```

d)  Change the x axis titles to "weight" and the title to "engine displacement".

```{r}
ggplot(mtcars, aes(x = wt, y = disp)) + 
  geom_point(color="blue") + 
  labs(x = "weight", title = "engine displacement", y = "")

# Alternative ways:
#
# ggplot(mtcars, aes(x = wt, y = disp)) + 
#   geom_point(color="blue") + 
#   xlab("weight") +
#   ylab("") +
#   ggtitle("engine displacement")
#   
# ggplot(mtcars, aes(x = wt, y = disp)) + 
#   geom_point(color="blue") + 
#   scale_x_continuous("weight") +
#   scale_y_continuous("") +
#   ggtitle("engine displacement")

```

e)  Change the `cyl` variable into a factor and add this as a color `aes` to the plot.

```{r}
mtcars %>% 
  mutate(cyl = factor(cyl)) %>% 
  ggplot(aes(x = wt, y = disp, color=cyl)) + 
  geom_point() + 
  labs(x = "weight", title = "engine displacement", y = "")
```

f)  Make a histogram of `wt` with `geom_histogram`, set binwidth on 0.5

```{r}
ggplot(mtcars, aes(x=wt)) + geom_histogram(binwidth = 0.5)
```

g)  Create multiple charts with `facet_wrap` for `cyl`

```{r}
ggplot(mtcars, aes(x=wt)) + geom_histogram(binwidth = 0.5) + facet_wrap(~cyl)
```

### Exercise 2

a)  Reproduce the following boxplot with `mtcars`. Use the functions `ggplot` and `geom_boxplot` from package `ggplot2`. NB: change `gear` into a factor. What happens if you do not do that? What is the difference between `color` and `fill`?

```{r, eval = TRUE}
library(dplyr)
library(ggplot2)
mtcars$gear <- as.factor(mtcars$gear)

ggplot(mtcars) +
    geom_boxplot(aes(x = gear, y = mpg, fill = gear)) +
  scale_fill_discrete("GEAR")
```

b)  Create the chart with box plots and add a scatter plot of `mpg` vs `gear`. Is it better to use `geom_point` or `geom_jitter`? Explain

```{r , eval=params$answers}
ggplot(mtcars, aes(x = gear, y = mpg, fill = gear)) +
    geom_boxplot() +
    geom_point()

ggplot(mtcars) +
    geom_boxplot(aes(x = gear, y = mpg, fill = gear)) +
    geom_jitter(aes(x = gear, y = mpg), width = 0.05, height = 0.0)
```

c)  Add information on the number of cylinders. Reproduce the following:

```{r ,  eval=TRUE}
mtcars <- mutate(mtcars, cyl = factor(cyl))

ggplot(mtcars) +
    geom_boxplot(aes(x = gear, y = mpg)) +
    geom_jitter(aes(x = gear, y = mpg, color = cyl), width = 0.1)
```

### Exercise 3

a)  Read the file *UnitedNations.txt*. Create a scatter plot with `tfr` vs life expectancy for women `lifeFemale`. Connect the aesthetic color to `region`.

```{r }
library(ggplot2)
library(dplyr)
library(readr)

UN <- read_delim(
    file = "./data/UnitedNations.txt",
    delim = " ",
    col_names = TRUE)

ggplot(data = UN) +
    geom_point(aes(x = lifeFemale, y = tfr, color = region)) 
```

b)  Add with `geom_smooth` a linear regression to the chart. Use the option `method="lm"` First add `x` and `y` and later on `color` What happens when you add `color`?

```{r, eval=params$answers}
ggplot(data = UN) +
    geom_point(aes(x = lifeFemale, y = tfr, color = region)) +
    geom_smooth(aes(x = lifeFemale, y = tfr), method = "lm")

ggplot(data = UN) +
    geom_point(aes(x = lifeFemale, y = tfr, color = region)) +
    geom_smooth(aes(x = lifeFemale, y = tfr, color = region), method = "lm")
```

c)  We will add GDP per capita in the graph. Connect aesthetic `size` to this variable, but since this variable is skewed, do a log transform. This can either be done with: `log` beforehand, or using the `scale_size` function. Experiment with option range to explore the sizes of the symbols.

```{r ,  eval=params$answers}
ggplot(data = UN) +
    geom_point(aes(
        x = lifeFemale,
        y = tfr,
        color = region,
        size = GDPperCapita)) 

ggplot(data = UN) +
    geom_point(aes(
        x = lifeFemale,
        y = tfr,
        color = region,
        size = GDPperCapita)) +
    
    scale_size(trans = "log", range = c(0, 3))
```

d)  Create with `ggplot` and `geom_histogram` a histogram of the values of `lifeFemale`. Change `geom_histogram` into a `geom_freqpoly` and connect `color` with `region`. Experiment with the `binwidth`.

```{r ,  eval=params$answers}
ggplot(data = UN) +
    geom_histogram(aes(x = lifeFemale, fill = region), bins = 25) + facet_wrap(~region)

ggplot(data = UN) +
    geom_histogram(aes(x = lifeFemale, fill = region), bins = 30) + facet_wrap(~region)

ggplot(data = UN) +
    geom_histogram(aes(x = lifeFemale, fill = region), bins = 35) + facet_wrap(~region)
```

(e) Create with `ggplot` en `geom_bar` a bar chart of the number of countries of the continents.

```{r}
ggplot(data = UN) +
    geom_bar(aes(x = region, fill = region))
```

### Exercise 4

a)  The package `ggplot2` contains the dataset `diamonds`. Create a histogram of the prices of diamonds. Add labels to the chart and axes

```{r ,  eval=params$answers}
library(ggplot2)

ggplot(data = diamonds) +
    geom_histogram(aes(x = price), bins = 50) +
    labs(
        tile = "Price of diamonds",
        x = "in $",
        y = "Count")
```

b)  Investigate the how `cut` and `price` interact.

- Create for each category a boxplot.
- Do the same with violin plot.

What do you notice?

```{r ,  eval=params$answers}
ggplot(data = diamonds) +
    geom_boxplot(aes(x = cut, y = price))

ggplot(data = diamonds) +
    geom_violin(aes(x = cut, y = price))
```

c)  Investigate the how `carat` and `price` interact.

-   Create a `geom_bin2d` color map and with add `geom_point` a scatterplot. Make sure the dots are white and 5% transparent. Any remarks on the relationship between these two variables?

```{r ,  eval=params$answers}
ggplot(data = diamonds) +
    geom_bin2d(aes(x = carat, y = price)) +
    geom_point(aes(x = carat, y = price), color = "white", alpha = 0.05, size = 0.1)
```

d)  Check you assumption: how do quality and carat relate? Explore/show this with a ggplot chart

```{r ,  eval=params$answers}
ggplot(data = diamonds) +
    geom_boxplot(aes(x = cut, y = carat))

ggplot(data = diamonds) +
        geom_violin(aes(x = cut, y = carat))
```

### Exercise 5

a)  Read the file *UnitedNations.txt* . Create a histogram of total fertility rate `tfr`.

```{r , eval=TRUE}
library(ggplot2)
library(dplyr)
library(readr)

UN <- read_delim(
    file = "./data/UnitedNations.txt",
    delim = " ",
    col_names = TRUE)
```

```{r , eval=params$answers}
ggplot(data = UN) +
    geom_histogram(aes(x = tfr), binwidth = .2)
```

b)  Which countries do not have a known `tfr`?

```{r , eval=params$answers}
summary(UN$tfr)
filter(UN, is.na(tfr))

UN <- filter(UN, ! is.na(UN$tfr))
```

c)  Use `geom_freqpoly`, `geom_boxplot` and `facet_wrap` to show the difference between the continents. Which graph is most clear?

```{r , eval=params$answers}

ggplot(data = UN) +
    geom_freqpoly(aes(x = tfr, col = region), bins = 20, size = 1)

ggplot(data = UN) +
    geom_histogram(aes(x = tfr), bins = 20) +
    facet_grid(region ~ .)

ggplot(data = UN) +
    geom_boxplot(aes(x = region, y = tfr))
```

e)  Use `facet_grid` and `geom_points` to create a plot with child mortality and `tfr` Add a regression line.

```{r , eval=params$answers}
ggplot(data = UN, aes(x = infantMortality, y = tfr)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_grid(region ~ .)
```

f)  Reproduce the following chart:

```{r, eval=TRUE}
UN <- mutate(
    UN,
    GDPperCapitaGroup = cut_number(GDPperCapita, 5))

selection <- filter(UN, ! is.na(GDPperCapita))
ggplot(data = selection, aes(x = GDPperCapitaGroup, y = tfr)) +
    geom_boxplot() +
    facet_grid(region ~ .)
```
