---
title: "Data management met R"
author: "ESTP Use of R in Official Statistics"
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
    fig_caption: false
editor_options: 
  chunk_output_type: console
---

```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment=NA)
library(tidyverse)
```

## Contents

- R packages and `tidyverse`
- Data set manipulation
    - Grammar of data manipulation
    - Aggregating
    - Data pipelines
    - Joining datasets
    - Restructuring dataset
- Reading
    - file names
    - csv, excel, spss, other
    - date/time variables

## R packages and tidyverse

### R packages

- Collection of standard and documented functions
- Can be published (after strict technical checks) on [CRAN](https://cran.r-project.org)

### Tidyverse

- A collection of packages meant for data-manipulation
- Consistent interface: data.frame in, data.frame out
- Website: [tidyverse.org](https://tidyverse.org)
- Focusses on data manipulation, not everything!


# Manipulation of datasets

## Data-manipulation

_factoid_: approx 80\% of time is datamanipulation...

### Reasons?
- Selecting subset of data (rows, columns)
- Deduplication (removing duplicate observations)
- Deriving/calculating new variables. 
- Counting/averaging, aggregating data
- Joining data sources
- Adapt structure
- Missing values
- Data contains outliers
- Data contains measurement errors / inconsistencies. 

## Grammar of data manipulation

### Concept

- Set of basic manipulations (_verbs_) that all have a data.frame (tibble) as input and
output
- Therefor these _verbs_ can be chained in any particular combination
- Manipulations for: 
    - data selection
    - deriving new variables 
    - aggregation
    - combining datasets (join)

### remarks
- This idea (algebraic closure) is same as  SQL
- Not all manipulation fit in the grammar

## The tidyverse datatype: `tibble`

- Nicer printing than `data.frame`
- With a `data.frame` `df` `df[,1]` returns a vector; with a `tibble`
  `tb` `tb[,1]` returns a tibble (with 1 column).
- The contain extra metadata use full for dplyr
- A tibble does not have to be a `data.frame` but also can refer to a table in a 
database (`dbplyr`!)

```{r}
tibble(x=1:3,y=letters[1:3])
```

## Implementation grammar of data manipulation: `dplyr` 

Offers a set of verbs for datamanipulations,
E.g:

\begin{tabular}{ll}
\code{filter}    & select rows\\
\code{select}    & select columns\\
\code{rename}    & rename columns\\
\code{distinct}  & select unique rows\\
\code{arrange}   & sort rows\\
\code{transmute} & caculate new columns\\
\code{mutate}    & add new columns
\end{tabular}





# Aggregating

## Aggregating

Aggregating is (arithmically) summarizing the data. Often this is done per group.
E.g. Calculate the mean income per educution level, age and gender.

### dplyr functions

\begin{tabular}{lp{0.7\textwidth}}
\code{group\_by} & Label a \code{tibble} so that each following operation is per group\\
\code{ungroup}   & Remove group labels\\
\code{summarize} & Calculate one (or more) aggregates.
\end{tabular}

## Aggregating with `dplyr`

```{r, eval=FALSE}
summarize(.data, ...)
```
With \code{...} you can give \code{<name>=<expression>} to calculate a statistic.

```{r, eval=TRUE}
summarize(iris,
          min_sl = min(Sepal.Length), # calculate min 
          max_sl = max(Sepal.Length)
          )
```

## Grouped aggregation with `dplyr`

```{r, eval=FALSE}
group_by(.data, ...)
```

With \code{...} you indicate which  (categorical) variable(s) are used for the groups

```{r}
summarize(group_by(iris, Species),
          min_sl = min(Sepal.Length),
          max_sl = max(Sepal.Length)) 
```

## Tabulating with `dplyr`

With the function `n()` you can count the number of rows. 

```{r}
summarize(group_by(iris, Species),
         number=n()
         )
```

## Tabulating with `dplyr` (2)

A shorter version is `count`  (same result)

```{r}
count(iris, Species, name="number")
```


## Pipelines

The pipe operator `|>` passes the output of the left side through as the first argument
of the following function. So:

```{r, eval=FALSE}
iris |>
  group_by(Species) |>
  summarize(min_sl = min(Sepal.Length),
            max_ls = max(Sepal.Length))
```

- The `|>` operator has been introduced in 2021 (as of R 4.1).
- Alternatively, the `%>%` operator (from the `magrittr` package, also in `tidyverse`) can be used.
- In RStudio, press CTRL + SHIFT + M to type this symbol.



## Pipelines (2)

A placeholder for the left side is the `_` symbol (and `.` for the `%>%` operator). 

```{r, eval=FALSE}
iris |>
  group_by(.data = _, 
           Species) |>
  summarize(.data = _,
            min_sl = min(Sepal.Length),
            max_ls=max(Sepal.Length))
```


```{r, eval=FALSE}
iris %>%
  group_by(.data  = ., 
           Species) %>%
  summarize(.data = .,
            min_sl = min(Sepal.Length),
            max_ls=max(Sepal.Length))
```

## Tabulating (1)

With `base` R function `table` you can cross tabulate. Create a matrix of counts.

```{r, eval=FALSE}
table(..., useNA = "no")
```
With \code{...} you supply the columns preferably in the form \code{<name>=<column>},
with \code{useNA} you indicate if `NA` should be counted.

```{r}
k <- c("1e","1e","3e","2e","2e")
table(k)
```

## Tabulating (2)

Multiple columns
```{r}
k <- c("1e","1e","3e","2e","2e")
g <- c("m","f","m","m","f")
table(class = k, gender = g)
```


## Tabulating (3)

The optie \code{useNA} has three possible values:

- \code{"no"}: records with NA are not counted
- \code{"ifany"}: for columns in which occurs `<NA>` is added to the matrix
- \code{"always"}: always count the number of NA's.


# Combining datasets

## Combine datasets with `dplyr`: stacking

```{r, eval=FALSE}
bind_rows(... ,.id=FALSE)
```

 \code{...} are a number of datasets (\code{data.frame}, \code{tbl}) to stack:
 the number of rows is the sum of the number of rows of the supplied data.frames.
 Columns are matched on name. Missing columns are filled with NA. Using `id` you 
 can recognize the origin of the data.

```{r,eval=FALSE}
bind_rows(women, women[2:1], women[-1],.id="ID")
```


## Joining datasets

\scriptsize
```{r}
(d1 <- tibble(key = letters[1:4], X = 1:4))
(d2 <- tibble(key = letters[2:5], Y = 6:9))
```
\normalfont


## Joining datasets: `inner_join`

```{r}
inner_join(d1, d2)
```

## Joining datasets: `left_join`

```{r}
left_join(d1, d2)
```

## Joining datasets: `full_join`

```{r}
full_join(d1, d2)
```

## 'Joining' datasets: `anti_join` | `semi_join`

\scriptsize
```{r}
anti_join(d1,d2)  # select records in d1 that do not match in d2
```

```{r}
semi_join(d1, d2) # selecteer records in d1 that do match in d2
```
\normalfont


## Pivoting datasets `pivot_longer` en `pivot_wider`

```{r}
wide <- tibble(name = c("jan","pier","joris")
      , age = c(23, 18, 27)
      , shoesize=c(44, 41, 43))
wide
```

## Pivoting datasets `pivot_longer` en `pivot_wider`

From wide to long: `pivot_longer``
\scriptsize
```{r}
pivot_longer(wide, age:shoesize, names_to = "variable")
```
\normalfont

Or:

\scriptsize
```{r}
long <- pivot_longer(wide, -name, names_to="variable")
```
\normalfont

## Pivoting datasets `pivot_longer` en `pivot_wider`

From long to wide: `pivot_wider`
\scriptsize
```{r}
pivot_wider(long, names_from = variable)
```
\normalfont

# Data types missing values

## Missing values

- Missing values in R are indicated with `NA`.
- A calculation with  `NA` almost always leads to `NA`.

```{r}
x <- c(1,4,2,NA,6)
c( mean1 = mean(x), mean2 = mean(x, na.rm=TRUE) )
```

- All statistical functions have the option to remove `NA` with argument
  `na.rm=TRUE`.


## Conversion of text to date variables.

### Base R

If the format is _exactly_ known,  `strptime` from base R can be used.

```{r}
strptime("2017/01/08",format="%Y/%m/%d",tz="CET")
```

## Conversion with the `lubdridate` package

### Lubridate 

Often it is simpler to use the `lubridate` package 

```{r,message=FALSE}
library(lubridate)
ymd("2017/01/08",tz="CET")
```

For converting you only need to know the order of day, month and year
`lubridate` contains the following conversion functions:
```
      dmy myd ydm mdy dym ymd
```

## `lubridate` versus `strptime` for converting dates

### `lubridate`

- Is simple
- Can be a bit slower
- Simpler to use with different locale settings

### `strptime`

- Can also convert time (not only date)
- Only works when the format is exactly known.

## Some usefull functions in  `lubridate`

\begin{center}
\begin{tabular}{ll}
\textbf{function} & \textbf{description}\\
\code{year}, \code{month}, \code{day} & year, month, day of the month\\
\code{wday} & day of the week\\
\code{week}, \code{isoweek} & week number, weeknr in ISO std.
\end{tabular}
\end{center}

```{r}
d <- ymd("2017-01-08")
c(USAweek = week(d), EUweek = isoweek(d))
```

## From date (`POSIXct`) to text

```{r,echo=FALSE, results=FALSE}
Sys.setlocale("LC_TIME","nl_NL.utf8")
```

Gebruik `format`
```{r}
d <- dmy("het was 2 januari 2016","8 jun 1996")
format(d, format="%A, %d %B '%y")
```

### Conversioncodes

Zie `?strptime` for a list of conversion codes

## Background

### `POSIXct`

This format store time as the number of seconds since `1970-01-01 00:00:00`.
During printing (to text) it is turned into a date time depending on the `locale` settings:
e.g.

- Calendar type (Gregorian, Chinese, Islamic, Coptic,...)
- Leap years (divisbable by 4, not by 100)
- Time zone (depends on region)
- Daylight Savings Time (idem)
- Leap seconds.

# I/O


## File names  in R

- File names are always between quotes.
- The directory separator is _always_ the forward slash

```{r,eval=FALSE}
dat <- read.csv("C:/users/joe/documents/foo.csv")
```


## Reading text files with `readr`

### Advantages
- faster than base R
- Specify columntypes
- Percentages are converted correctly
- Good error reporting

### Disadvantages
- Less robust

## Reading csv files with `readr`

```{r,echo=FALSE}
library(readr)
```

```{r,eval=FALSE}
library(readr)
mydata <- read_csv(file)
# or
mydata <- read_csv2(file)
```

Specify column type

```{r}
write_csv(iris,"myiris.csv")
dat <- read_csv("myiris.csv", col_types = "ddddc")
```

## Compact column types

\begin{center}
\begin{tabular}{ll}
\textbf{Code} & \textbf{Type}\\
\code{l} & \code{logical}\\
\code{i} & \code{integer}\\
\code{d} & \code{double} (=\code{numeric})\\
\code{c} & \code{character} (text)\\
\code{D} & \code{datum} (in locale format)\\
\code{T} & \code{datum/tijd} (in ISO8601) format.\\
\code{\_} of \code{-} & Skip column
\end{tabular}
\end{center}


## When reading does not succeeed 

e.g. "money", "kg" etc.

- Read in as \code{character}
- Use \code{parse\_number} to extra the numbers.

```{r}
x <- c("10%", "EUR 7,-", "$12")
x
parse_number(x)
```

## Other file formats

### The `haven` package

\begin{tabular}{ll}
SPSS  & \code{read\_spss}, \code{read\_por}\\
STATA & \code{read\_stata}\\
SAS   & \code{read\_sas}
\end{tabular}

### The `foreign` package (a standard R package)

Functions for reading \code{arff}, \code{dbf},
\code{dba}, \code{epiinfo}, \code{mtp}, \code{octave}, \code{systat},
and \code{xport} files.


## Databases and high performance computing

### Databases

`dbplyr` provides a `dplyr` back-end for databases: see demo in the next slides

### High performance computing

`sparklyr` provides a `dplyr` back-end for Spark, an engine for high performance computing (discussed at ESTP course Big Data Tools for Advanced Users)


## `dbplyr`, database back-end (demo 1/8)

\scriptsize

```{r}
library(dplyr)
library(dbplyr)

# create a database
con <- DBI::dbConnect(RSQLite::SQLite())

# copy some data to the database
DBI::dbWriteTable(con, "iris", iris)

# retrieve a reference to the table in the database
tbl_iris <- tbl(con, "iris")
```

## `dbplyr`, database back-end (demo 2/8)


\scriptsize

```{r}
# this now a link to the table in the database
tbl_iris
```

## `dbplyr`, database back-end (demo 3/8)

\scriptsize

```{r}
query <- 
  tbl_iris %>% 
  mutate(silly = 1)

print(query)
```

## `dbplyr`, database back-end (demo 4/8)

\scriptsize

```{r}
show_query(query)
```

## `dbplyr`, database back-end (demo 5/8)

\scriptsize

```{r}
# this now a link to the table in the database
query <- 
  tbl_iris %>% 
  group_by(Species) %>% 
  summarize( mean_sepal_length = mean(Sepal.Length, na.rm=TRUE)
           , mean_sepal_width = mean(Sepal.Width, na.rm = TRUE)
           ) %>% 
  mutate(ratio = mean_sepal_length / mean_sepal_width)

print(query)
```

## `dbplyr`, database back-end (demo 6/8)

\scriptsize

```{r}
show_query(query)
```

## `dbplyr`, database back-end (demo 7/8)

\scriptsize

```{r}
# you can also directly use SQL on the database
DBI::dbGetQuery(con, "SELECT * FROM iris")
```

## `dbplyr`, database back-end (demo 8/8)

### Close connection

```{r}
DBI::dbDisconnect(con)
```

