---
title: "Plotting with ggplot2"
author: "ESTP Use of R in Official Statistics"
output: 
  beamer_presentation:
    df_print: kable
    includes:
      in_header: ../header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  out.width = "70%",
  fig.showtext = TRUE,
  fig.align = "center"
)
library(ggplot2)
library(dplyr)
```

## Graphs

> When communicating results to non-technical types there is nothing better than a clear visualization to make your point.

> Numerical quantities focus on expected values, graphical summaries on unexpected values.

__John Tukey__


## Role visualization in  data-analysis

```{r, echo=FALSE}
knitr::include_graphics("fig/model-slide.pdf")
```


## Plotting in R

R has great plotting facilities:

- Very fine-grained control over plot
- Most functionality is/was focused on static images.

##


- `base`/`graphics` offers extended possibilities for making and tweaking graphics
- `ggplot2` creates by default good graphs
- Lots of extra packages for making specialized graphs.
- E.g.: `highchartR`,   `plotly`, `ggvis`, `dyGraph`.

## Exporting figures/graph

### By hand (not recommended):

- In Rstudio by default the plotting panel is used.
- You can export plot from RStudio using the menu: "Export".

### Setting programmatically a device:

1. Set as plotting device:

    - `pdf` Creates a  pdf device (document)
    - `png` Creates a png device (document)
    - ...
2. Use your plotting commands: plots will be writting to disk.
3. Close the device: `dev.off()`


`ggplot` also has `ggsave`

## Grammar of graphics


A systematic formulation ("grammar") of how charts are build:

- Data
- Data mapping (which variables are used)
- Geometry: points, lines, ...?
- Statistics: identity, binned, density, ...?
- Scale: linear/logarithmic, color scales
- Coordinates: Cartesian or polar
- Facets (small multiples)

## ggplot2

- Grammar of Graphics plotting library
- Hadley Wickham

```R
install.packages("ggplot2")
```

## Example data

\scriptsize

```{r, echo=TRUE}
head(mtcars)
```

## Example data

\scriptsize

```{r}
str(mtcars)
```

## Definition of statistical graphic

> A statistical graph is a __mapping__ of __variables__ on __aesthetics__ of
__geometric__ objects in which
variable values are __scaled__ to __aesthetic values__.

- geom_<g>
- scale_<a>

## geometric objects?

- line, bar, area, polygon, etc.

In `ggplot` these are combined into a *geom_*etric object


## Graphs with ggplot2

 - Gebaseerd op *Grammar of Graphics*;
 - Graph = Data + Mapping + Geometry + Coördinate system;
 - Graph as one or more **geom**etries.
 - Geometry has **aes**thetic features (size, position, color etc.);
 - Columns are mapped to aesthetics.
 
## Example scatterplot

::::::{.columns}

:::{.column}
\scriptsize
```{r scatter, echo=TRUE, eval=FALSE}
library(ggplot2)
ggplot( data = mpg
      , aes(x = displ, y = hwy)) +
    geom_point()
```
:::

:::{.column}
```{r scatter, eval=TRUE, echo=FALSE, out.width="95%"}
```

:::

::::::

## scatter plot (color)

::::::{.columns}

:::{.column}
\scriptsize
```{r scattercol, echo=TRUE, eval=FALSE}
mtcars$cyl <- factor(mtcars$cyl)
library(ggplot2)
ggplot( data = mpg
      , aes( x = displ
           , y = hwy
           , color = cyl
           )) +
    geom_point()
```
:::

:::{.column}
```{r scattercol, eval=TRUE, echo=FALSE, out.width="95%"}
```

:::

::::::

## scatter plot (shape)

\tiny
```{r}
mtcars$cyl <- factor(mtcars$cyl) 
ggplot(data = mtcars, aes(x=disp, y=hp, shape=cyl)) + geom_point()
```

## scatter plot (shape)

\tiny
```{r}
mtcars$cyl <- factor(mtcars$cyl) 
ggplot(data = mtcars, aes(x=disp, y=hp, shape=cyl, color = cyl)) + geom_point()
```

## Assignment

- Create a histogram from `carat` column in the `diamonds` dataset (with `geom_histogram`)

- Play around with the `binwidth`: what is the story of the data?

## Facets

Split any ggplot plot on a categorical variable into small multiples

- Very useful during analyses!

- By default same scales, to make comparison easy

## Facets 

\tiny
```{r}
ggplot(mtcars) + geom_histogram(aes(x = disp), binwidth = 25) + facet_wrap(~cyl)
```

## Facets (2)

\tiny
```{r}
mtcars$cyl <- factor(mtcars$cyl) 
ggplot(data = mtcars, aes(x=disp, y=hp)) + geom_point() + facet_wrap(~cyl)
```

##

 - Specify data with  `ggplot`;
 - add extra layers or properties using a `+`
 - Specify a geometry with `geom_`.
 - Map a column to an aesthetic with `aes`;
 - Specify a geometry with `geom_`.
 - Each aesthetic has a `scale_`
 - See cheat sheet.
 

## ggplot2 tricks

- Each plot can be stored in a variable and be adjust

```{r, eval = FALSE}
p <- ggplot(mtcars) + geom_point(aes(x = disp, y = hp))
p + facet_wrap(~cyl)
p + facet_wrap(~am)
```

- A plot can be save easily with `ggsave`

```{r, eval = FALSE}
ggsave("test.pdf") # remembers last plot
# or when assigned to variable
ggsave("test.png", plot = p)
```


## ggplot2 useful packages

- `ggforce`, annotation

- `gganimate` create an animation from a ggplot 

- `ggrepel`, nice label positioning of annotated labels.

- `plotly`, create interactive charts (function `ggplotly`)

